Confronting preferential sampling when analyzing population distributions: diagnosis and model-based triage \footnotemark[2]

### 2017-04-24 manuscript/AppendixS2.tex
Restructured R files for package compilation and public consumption (including putting TMB .cpp code in the src directory for automatic compilation when package is loaded). Final changes to ms. before submission.
This commit corresponds to code and manuscript status when the manuscript was accepted at MEE. It represents the first commit where cross validation was used in the sim example.

\vspace{0.7cm}
Paul B. Conn$^{1,*}$, James T. Thorson$^2$, Devin S. Johnson$^1$
\end{center}
\vspace{0.5cm}

@@

In addition to the models reported in the paper, we attempted to fit two different models with a trend surface specification for the preferential sampling process.  The models differed by whether or not predictive covariates were also modeled (the ice covariates described in the main article text), but otherwise the preferential sampling process was modeled similarly.  Specifically, when the probability $R_i$ that a location $i$ is selected for sampling is modeled as
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(\textrm{expit}(\nu_i))),
\end{eqnarray}
where
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i,
\label{eq:nu}
\end{eqnarray}
we implemented a trend surface model by writing $\bf{B}$ as
\begin{eqnarray}
  \label{eq:B}
  \textbf{B} & = & {\bf C}{\bf b}.
\end{eqnarray} 
Here, ${\bf b}$ is a column vector consisting of three estimated parameters, $b_0$, $b_1$, and $b_2$, and ${\bf C}$ is an $(S \times 3)$ design matrix constructed as follows:
\begin{eqnarray}
  \textbf{C} & = & \left[ 
    \begin{array}{ccc}
    1 & northing_1 & easting_1 \\
    1 & northing_2 & easting_2 \\
    \vdots & \vdots & \vdots \\
    1 & northing_S & easting_S 
    \end{array} 
  \right].
\end{eqnarray}
The northing and easting values are simply the $y$ and $x$ coordinates associated with each grid cell centroid on the particular geographical projection of the earth's surface that we were using (in our case, EASE Grid 2.0).

\hspace{0.5in} On the surface, there did not appear to be any numerical errors when fitting these trend surface models to bearded seal data.  The Hessian was invertible, and incorporation of the trend surface model resulted in a much lower AIC score than the other models (AIC scores were 19 and 26 points better than the highest ranked AIC model reported in the paper for the trend surface models omitting and including covariate effects, respectively).  However, when examining estimates graphically, it was clear that the trend surface models suffered from some parameter identification and/or aliasing issues (Fig. \ref{fig:bearded_RE_maps}).  Specifically, most variation in the probability of sampling was passed on from the $\boldsymbol{\eta}$ random effect to the shared random effect $\boldsymbol{\delta}$ such that it was difficult to disentangle the sampling process from the abundance process.  To our mind, this was a clear sign that the model was overparameterized, leading to poor estimation.

>>>>

\rm
\vspace{0.7cm}
Paul B. Conn$^{1,*}$, James T. Thorson$^2$, Devin S. Johnson$^1$
\end{center}
\vspace{0.5cm}

@@

We attempted to fit four models that estimated preferential sampling parameter(s) to bearded seal data.  The models differed by whether or not habitat and landscape covariates were modeled, and how the preferential sampling effect was modeled.  Specifically, when the probability $R_i$ that a location $i$ is selected for sampling is modeled as
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(\textrm{logit}^{-1}(\nu_i)),
\end{eqnarray}
where
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i,
\label{eq:nu}
\end{eqnarray}
we considered two formulations for \textbf{B}.  In the first, it was simply set to $b \textbf{I}$, where $\textbf{I}$ is an $(S \times S)$ identity matrix and $b$ is an estimated parameter.  In the second, we implemented a trend surface model by writing $\bf{B}$ as
\begin{eqnarray}
  \label{eq:B}
  \textbf{B} & = & {\bf C}{\bf b} \textbf{I}.
\end{eqnarray}
Here, ${\bf b}$ is a column vector consisting of three estimated parameters, $b_0$, $b_1$, and $b_2$, and ${\bf C}$ is an $(S \times 3)$ design matrix constructed as follows:
\begin{eqnarray}
  \textbf{C} & = & \left[
    \begin{array}{ccc}
    1 & northing_1 & easting_1 \\
    1 & northing_2 & easting_2 \\
    \vdots & \vdots & \vdots \\
    1 & northing_S & easting_S
    \end{array}
  \right].
\end{eqnarray}
The northing and easting values are simply the $y$ and $x$ coordinates associated with each grid cell centroid on the particular geographical projection of the earth's surface that we were using (in our case, EASE Grid 2.0).

\hspace{0.5in} On the surface, there did not appear to be any numerical errors when fitting preferential sampling models to bearded seal data.  The Hessian was invertible, and incorporation of the preferential sampling effects resulted in much lower AIC scores than the other models (AIC scores were close to 100 points better than the highest ranked AIC model omitting preferential sampling effects).  However, when examining estimates graphically, it was clear that the preferential sampling models suffered from some aliasing issues (Fig. \ref{fig:bearded_RE_maps}).  Specifically, most variation in the probability of sampling was passed on from the $\boldsymbol{\eta}$ random effect to the shared random effect $\boldsymbol{\delta}$ such that it was difficult to disentangle the sampling process from the abundance process.  To our mind, this was a clear sign that the model was overparameterized, leading to poor estimation.
